Scaling rates of true polar wander in convecting planets and moons

### 2017-04-17 tpw_rate.tex
Spelling.
Edits from Bruce.

\begin{abstract}
Mass redistribution in the convecting mantle of a planet can cause perturbations in its moment of inertia tensor. 
Conservation of angular momentum dictates that these perturbations can change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). 
Although the existence of TPW on Earth is well-verified, its rate and magnitude over geologic time scales remain controversial. 
Here we present scaling analyses and numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors. 
For simple rotating convection, the most important parameters are the Rayleigh number, the rotation rate, and the size of relative density fluctuations (i.e. thermal expansivity times the temperature variations). 
We identify timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander. 
These timescales, as well as the relative sizes of convective anomalies, control the rate and magnitude of TPW.
This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and when they are likely to occur.
Finally, we discuss implications for large-scale TPW in Earth's past, which has been suggested to be important for life and climate history.
\end{abstract}

@@

Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle,  which we support with numerical simulations.

@@

\begin{equation}
\frac{d \mathbf{H}}{dt} + \mitbf{\Omega} \times \mathbf{H} = 0
\label{eq:euler}
\end{equation}
where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity vector.
On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
In this case, Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
For a slowly convecting fluid, such as a planetary mantle, the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
\begin{equation}
\mitbf{\Omega } (t)\times ( \mathbf{I}(t) \cdot \mitbf{\Omega }(t) ) = 0.
\label{eq:quasistatic_liouville}
\end{equation}
Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for $\mathbf{I}(t)$, where the eigenvectors correspond to the principal axes and the eigenvalues correspond to the principal moments (where the most stable orientation of the planet corresponds to rotating about the largest principal axis).
In practice, this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
The moment of inertia tensor in Equation~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and surface deflections due to density anomalies.
Here we are interested in contributions from mantle convection, so we restrict our attention to these processes.

@@

\subsection{A note on reference frames}
\label{sec:reference_frames}
True polar wander can be described in different reference frames, and this choice is fundamentally an arbitrary one.
However, certain aspects of the physics can be made much simpler by an appropriate choice of the reference frame.
In our treatment of TPW, we will refer to three different reference frames:
\begin{itemize}
\item First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
\item Second, there is the body-fixed (or geographic) frame.
By definition, the rotation of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
A terrestrial no-net-rotation or hotspot reference frame are common choices for the body-fixed frame.
Treatments of gravitational or rotational deformation of a planet are most naturally expressed in the body-fixed frame,
as described in Section~\ref{sec:rotational_deformation}.
\item Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
which we will refer to as the ``$\mathbf{E}$-frame.'' 
Redistribution of mantle mass anomalies due to convection changes the principal axes of $\mathbf{E}$, 
causing the $\mathbf{E}$-frame to slowly rotate with respect to the geographic frame. We denote this drift 
by a rotation vector $\mitbf{\Psi}$ (see Section~\ref{sec:theta}).
\end{itemize}

These reference frames and vectors are illustrated in Figure~\ref{fig:reference_frames}.

@@

The part of the moment of inertia due to the elastic rotational deformation in the body-fixed frame 
is traditionally related to the degree-two part of the gravity field via MacCullagh's formula \citep{munk1960rotation}:
\begin{equation}
J_{ij} = \frac{k a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right)
\label{eq:elastic_deformation}
\end{equation}
where $k$ is an elastic Love number, $a$ is the semimajor axis of the planet, and $G$ is the gravitational constant.
This result may be extended to a viscoelastic rheology via the viscoelastic correspondence principle \citep[e.g.][]{peltier1974impulse}:
\begin{equation}
J_{ij} = \frac{k(t) a^5}{3 G} * \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right)
\label{eq:viscoelastic_deformation}
\end{equation}
where $k$ is now a time-dependent viscoelastic Love number which is convolved with the time-dependent rotation vector.

The infinite-time limit of Equation~\eqref{eq:viscoelastic_deformation} for a 
constant rotation vector around the $z$-axis implies
\begin{equation}
\begin{aligned}
J_{zz} = C = &\frac{2}{3} \frac{k_f a^5 \Omega^2}{3 G} \\
J_{xx} = J_{yy} = A = -&\frac{1}{3} \frac{k_f a^5 \Omega^2}{3 G} \\
\end{aligned}
\end{equation}
where $C$ and $A$ are the polar and equatorial moments of inertia, respectively, 
and $k_f$ is the fluid limit of $k$. 
We can solve for $k_f$ in terms of $C-A$:
\begin{equation}
k_f = \frac{3 G (C-A)}{\Omega^2 a^5}.
\label{eq:fluid_love}
\end{equation}
This fluid-limit representation of the deformation does not allow for any disequilibrium between $J_{ij}$ and $E_{ij}$,
so it does not permit TPW.

\citet{ricard1993polar} obtain an approximation to Equation~\eqref{eq:viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.  
This introduces a new parameter, termed $T_1$, which can be seen as a weighted relaxation time for the system.  
This simple approximation in the Laplace domain allows for an analytical transformation back into the time domain, and neglecting second order terms in $\dot{\Omega}$ we find:
\begin{equation}
J_{ij} = \frac{k_f a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right) -
 \frac{k_f a^5 T_1}{3G} \left(\dot{\Omega}_i \Omega_j + \Omega_i \dot{\Omega}_j - \frac{2}{3} \Omega_q \dot{\Omega}_q \delta_{ij} \right).
\label{eq:rotational_deformation}
\end{equation}
The two terms of this equation have simple interpretations.  
The first term corresponds to the fluid limit of rotational deformation (in the absence of any long-term elastic strength).  
The second term represents the lag in the moment of inertia due to the viscous adjustment of the rotational bulge, where $T_1$ is the characteristic time constant for this adjustment.
Since the first term represents the fluid limit of rotational deformation, it automatically satisfies Equation~\eqref{eq:quasistatic_liouville}, and hence does not contribute to the polar wander problem.

@@

The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
This part of the moment of inertia may also be parameterized using a viscoelastic Love number approach:
\begin{equation} 
\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
\end{equation}
where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, 
and $C_{ij}$ is the moment of inertia due solely to the internal load.
Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
\begin{equation}
E_{ij} = (1+k^L_f) C_{ij}.
\end{equation}
An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
permits more general treatments of mantle rheology.

@@

We are in a position to address the rates of true polar wander for a given convective moment $\mathbf{E}$.
A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
Substituting Equation~\eqref{eq:rotational_deformation} into Equation~\eqref{eq:disequilibrium} we find
\begin{equation}
\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right).
\label{eq:liouville}
\end{equation}
Introducing a unit vector $\mitbf{\omega} = \mitbf{\Omega}/\|\mitbf{\Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for $\dot{\mitbf{\omega}}$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right].
\label{eq:liouville_ode}
\end{equation}
Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
This correspondence permits useful insights (see below).
Let $\dot{\Theta} = \vert \dot{\mitbf{\omega}} \vert$ denote the rate of polar wander.
Evaluating the scalar product $\dot{\Theta}^2 = \dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}}$ gives
\begin{equation}
\dot{\Theta}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
\label{eq:polar_wander_rate_tensor}
\end{equation}

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/reference_frames.pdf}
\caption{Relevant vectors and angles for the TPW analysis. The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively. Since the choice of geographic axes is arbitrary, we assume that at this instant the $\mathbf{E}$-frame and the geographic frames are colocated (though at future times they will not be). The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $\mathbf{e}_3$-axis. For illustration, we assume the longitude $\phi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane. True polar wander moves the rotation axis towards the $\mathbf{e}_3$-axis. We denote this motion by $\dot{\mitbf{\omega}}$ in the geographic frame. However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$-frame. This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$, which is defined by colatitude $\beta$ and longitude $\gamma$. The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$-frame.  Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Sections~\ref{sec:reference_frames} and \ref{sec:theta}}
\label{fig:reference_frames}
\end{figure*}

Our goal is to quantify the polar wander rate $\dot{\Theta}$ due to a convective perturbation $\mathbf{E}$ in the moment of inertia. 
The rate defined by Equation~\eqref{eq:polar_wander_rate_tensor} is expressed in the body-fixed geographic frame, 
which is also the reference frame used to measure TPW. 
However, the physics of the right-hand-side of Equation~\eqref{eq:polar_wander_rate_tensor} 
is more naturally expressed in the reference frame of the principal axes of $\mathbf{E}$. 
In general the $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, so the time derivative of $\omega$ is different in these two frames.
However, Equation~\eqref{eq:polar_wander_rate_tensor} is a scalar equation, which means it is invariant to rotations.
Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of 
$\mitbf{\omega}$ in the $\mathbf{E}$-frame with colatitude $\theta$ and longitude $\phi$ (see Figure~\ref{fig:reference_frames}).
Plugging this description of $\mitbf{\omega}$ into Equation~\eqref{eq:polar_wander_rate_tensor},
and after some tedious algebra, we find
\ifdetail

And here is the tedious algebra.  Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$.
We are interested in the rate of true polar wander, which is the angular change of the pole
per unit time. We can transform Equation~\eqref{eq:liouville_ode} into a scalar equation by dotting it
with $\dot{\mitbf{\omega}}$, since we are interested in the magnitude of the rate, rather than
the direction at the moment.

\begin{equation}
\begin{aligned}
\dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}} &= 
A^2 \left[ \mathbf{E}\cdot \mitbf{\omega} - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)\mitbf{\omega}\right]
\cdot \left[ \mathbf{E}\cdot \mitbf{\omega} - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)\mitbf{\omega}\right] \\
\dot{\mitbf{\omega}}^2  &= 
A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)\cdot \left(\mathbf{E}\cdot \mitbf{\omega}\right)
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right) \left( \mitbf{\omega}\cdot \mathbf{E} \cdot \mitbf{\omega} \right) 
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right) \left( \mitbf{\omega}\cdot \mathbf{E} \cdot \mitbf{\omega} \right) 
 + \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 \mitbf{\omega}\cdot \mitbf{\omega} 
\right]\\
\dot{\mitbf{\omega}}^2  &= 
A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2.
\right]\\
\end{aligned}
\end{equation}
Since this is a scalar equation, we can arbitrarily choose a coordinate system in which to evaluate it.
It is most convenient to choose the principal axes of $\mathbf{E}$, where
\begin{equation}
\mathbf{E} = 
\begin{bmatrix}
\lambda_1 & 0 & 0 \\
0 & \lambda_2 & 0 \\
0 & 0 & \lambda_3 \\
\end{bmatrix}
\end{equation}
\begin{equation}
\mitbf{\omega} = 
\begin{bmatrix}
\sin{\theta} \cos{\phi} \\
\sin{\theta} \sin{\phi} \\
\cos{\theta} \\
\end{bmatrix}.
\end{equation}

As a warm-up, consider the case where $\phi = 0$, then  
\begin{equation}
\mitbf{\omega} = 
\begin{bmatrix}
\sin{\theta}  \\
0 \\
\cos{\theta} \\
\end{bmatrix}
\end{equation}
We can evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
\begin{equation}
\mathbf{E}\cdot \mitbf{\omega} = 
\begin{bmatrix}
\lambda_1 \sin{\theta} \\
0 \\
\lambda_3 \cos{\theta}
\end{bmatrix}
\end{equation}

\begin{equation}
\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
\end{equation}

\begin{equation}
\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
\lambda_1 \sin^2{\theta} + \lambda_3 \cos^2{\theta}
\end{equation}

\begin{equation}
\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
\lambda_1^2 \sin^4{\theta} + \lambda_3^2 \cos^4{\theta} + 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta}
\end{equation}
Therefore, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
\begin{equation}
\begin{aligned}
\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
 \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
 &= \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
 - \lambda_1^2 \sin^4{\theta} - \lambda_3^2 \cos^4{\theta} - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
 &= \lambda_1^2 \sin^2{\theta} (1 - \sin^2{\theta}) + \lambda_3^2 \cos^2{\theta} (1 - \cos^2{\theta})
  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
 &= \lambda_1^2 \sin^2{\theta} \cos^2{\theta} + \lambda_3^2 \cos^2{\theta} \sin^2{\theta}
  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
 &= (\lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \sin^2{\theta} \cos^2{\theta} \\
 &= \frac{1}{2} (\lambda_3 - \lambda_1)^2 \sin{2 \theta} \\
\end{aligned}
\end{equation}

Now consider the more general case where $\phi \ne 0$.
Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
\begin{equation}
\mathbf{E}\cdot \mitbf{\omega} = 
\begin{bmatrix}
\lambda_1 \sin{\theta} \cos{\phi}\\
\lambda_2 \sin{\theta} \sin{\phi} \\
\lambda_3 \cos{\theta}
\end{bmatrix}
\end{equation}

>>>>

\begin{abstract}
Mass redistribution in the convecting mantle of a planet causes perturbations in its moment of inertia tensor. 
Conservation of angular momentum dictates that these perturbations change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). 
Although the existence of TPW on Earth is firmly established, its rate and magnitude over geologic time scales remain controversial. 
Here we present scaling analyses and numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors. 
For simple rotating convection, we identify a set of dimensionless parameters that characterize true polar wander. We use these parameters to define  timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander. 
These timescales, as well as the relative sizes of convective anomalies, control the rate and magnitude of TPW.
This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and relates them to a  broader class of events that enable large and often rapid TPW. We expect these events to have been more frequent in Earth’s past.
\end{abstract}

@@

Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle. We identify a set of dimensionless parameters that fully characterize TPW and use numerical simulations to quantify the expected dependence on these parameters. Finally, we use the scaling results to make predictions about TPW over Earth's history and for other planetary bodies.

@@

\begin{equation}
\frac{d \mathbf{H}}{dt} + \mitbf{\Omega} \times \mathbf{H} = 0
\label{eq:euler}
\end{equation}
where $\mathbf{H} = \mathbf{I} \cdot \mitbf{\Omega}$ is the angular momentum vector, $\mathbf{I}$ is the moment of inertia tensor, and $\mitbf{\Omega}$ is the angular velocity vector.
On a dynamic planet $\mathbf{I}$ may be a function of time, so to conserve angular momentum $\mitbf{\Omega}$ must be also vary with time.
In this case, Equation~\eqref{eq:euler} is often called the Liouville equation \citep[e.g.][]{munk1960rotation}.
For a slowly convecting fluid, such as a planetary mantle, the inertial term $\partial \mathbf{H} / \partial t$ is negligible, so we may solve the simplified quasistatic equations
\begin{equation}
\mitbf{\Omega } (t)\times ( \mathbf{I}(t) \cdot \mitbf{\Omega }(t) ) = 0.
\label{eq:quasistatic_liouville}
\end{equation}
Equation~\eqref{eq:quasistatic_liouville} indicates that $\mitbf{\Omega}$ and $\mathbf{H}$ are parallel, so a solution for $\mitbf{\Omega}(t)$ is equivalent to solving an eigenvalue problem for the principle values of $\mathbf{I}(t)$;  the eigenvectors define the principal axes and the eigenvalues correspond to the principal moments of inertia. The rotation axis is expected to align with the axis for the largest moment.  
In practice, this eigenvalue approach has been often used in previous studies for computing the spin axis of a planet \citep[e.g.][]{steinberger1997changes, roberts2007cause}.
The moment of inertia tensor in Equation~\eqref{eq:quasistatic_liouville} includes all contributions to the mass structure
of the planet, including the spherically symmetric mass distribution, rotational deformation, deformation due to self gravity, internal and surface density anomalies, and the associated surface deflections.
Here we are interested in those contributions that arise from mantle convection.

@@

\subsection{A note on reference frames}
\label{sec:reference_frames}
True polar wander can be described in different reference frames, and this choice is fundamentally an arbitrary one.
However, certain aspects of the physics can be made much simpler by an appropriate choice of the reference frame.
In our treatment of TPW, we will refer to three different reference frames.  First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
 Second, there is the body-fixed (or geographic) frame.
By definition, the rotation of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
A terrestrial no-net-rotation or hotspot reference frame are common choices for the body-fixed frame.
Treatments of gravitational or rotational deformation of a planet are most naturally expressed in the body-fixed frame,
as described in Section~\ref{sec:rotational_deformation}.
 Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
which we will refer to as the ``$\mathbf{E}$-frame.'' 
Redistribution of mantle mass anomalies due to convection changes the principal axes of $\mathbf{E}$, 
causing the $\mathbf{E}$-frame to slowly rotate with respect to the geographic frame.

@@

Rotational deformation in an elastic body is traditionally related to the degree-two part of the 
centrifugal potential using a Love-number formalism. When the perturbation in gravitational
potential is related to the moments of inertia using MacCullagh's formula  \citep{munk1960rotation}, we find:
\begin{equation}
J_{ij} = \frac{k a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right)
\label{eq:elastic_deformation}
\end{equation}
where $k$ is an elastic Love number, $a$ is the semimajor axis of the planet, and $G$ is the gravitational constant. Both $\Omega_i$ and
$J_{ij}$ are defined in the body-fixed frame. An extension to  viscoelastic rheology is permitted by the viscoelastic correspondence principle \citep[e.g.][]{peltier1974impulse}:
\begin{equation}
J_{ij} = \frac{k(t) a^5}{3 G} * \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right)
\label{eq:viscoelastic_deformation}
\end{equation}
where $k$ is now a time-dependent viscoelastic Love number which is convolved with the time-dependent rotation vector.

The infinite-time limit of Equation~\eqref{eq:viscoelastic_deformation} for a 
constant rotation vector around the $z$-axis implies
\begin{equation}
\begin{aligned}
J_{zz} = C = &\frac{2}{3} \frac{k_f a^5 \Omega^2}{3 G} \\
J_{xx} = J_{yy} = A = -&\frac{1}{3} \frac{k_f a^5 \Omega^2}{3 G} \\
\end{aligned}
\end{equation}
where $C$ and $A$ are the polar and equatorial moments of inertia, respectively, 
and $k_f$ is the fluid limit of $k$. 
We can solve for $k_f$ in terms of $C-A$:
\begin{equation}
k_f = \frac{3 G (C-A)}{\Omega^2 a^5}.
\label{eq:fluid_love}
\end{equation}
This infinite-time limit  ensures that the left-hand side of Equation (4) vanishes. It follows that the contribution from the convective perturbation $E_{ij}$ also vanishes, so TPW is not permitted.

\citet{ricard1993polar} obtain an approximation to Equation~\eqref{eq:viscoelastic_deformation} which retains its long-time behavior by entering the Laplace domain and truncating a Taylor series for $k(s)$ to first order.  
This introduces a new parameter, $\tau_R$, which can be seen as a weighted relaxation time for the system.  
This simple approximation in the Laplace domain allows for an analytical transformation back into the time domain. Neglecting second order terms in $\dot{\Omega}$ we find:
\begin{equation}
J_{ij} = \frac{k_f a^5}{3 G} \left( \Omega_i \Omega_j - \frac{1}{3} \Omega_q \Omega_q \delta_{ij} \right) -
 \frac{k_f a^5 \tau_R}{3G} \left(\dot{\Omega}_i \Omega_j + \Omega_i \dot{\Omega}_j - \frac{2}{3} \Omega_q \dot{\Omega}_q \delta_{ij} \right).
\label{eq:rotational_deformation}
\end{equation}
The two terms of this equation have simple interpretations.  
The first term corresponds to the fluid limit of rotational deformation (in the absence of any long-term elastic strength).  
The second term represents the lag in the moment of inertia due to the viscous adjustment of the rotational bulge, where $\tau_R$ is the characteristic time constant for this adjustment.
Since the first term represents the hydrostatic limit of rotational deformation, it automatically satisfies Equation~\eqref{eq:quasistatic_liouville}, and hence does not contribute to the polar wander problem.

@@

The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
Both of these contributions may also be parameterized using a viscoelastic Love number approach:
\begin{equation} 
\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
\end{equation}
where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, 
and $C_{ij}$ is the moment of inertia due solely to the internal load.
When the timescale of the surface response is short compared to the true polar wander timescale, it is reasonable to  use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
\begin{equation}
E_{ij} = (1+k^L_f) C_{ij}.
\end{equation}
An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
permits more general treatments of mantle rheology.

@@

We are in a position to address the rates of true polar wander for a given convective moment $\mathbf{E}$.
A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
Substituting Equation~\eqref{eq:rotational_deformation} into Equation~\eqref{eq:disequilibrium} we find
\begin{equation}
\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right).
\label{eq:liouville}
\end{equation}
Introducing a unit vector $\mitbf{\omega} = \mitbf{\Omega}/\|\mitbf{\Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for $\dot{\mitbf{\omega}}$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right].
\label{eq:liouville_ode}
\end{equation}
In order to evaluate $\dot{\mitbf{\omega}}$ we require knowledge of both $\mathbf{E}$ and $\mitbf{\omega}$ in the geographic frame (i.e. the frame used to derive Equation (13)). The same frame is also used to measure TPW, but the expressions on the right-hand-side of Equation (13) are more naturally evaluated in the reference frame of the principal axes of $\mathbf{E}$.
We can avoid the need to reconcile these two frames by confining our attention to the rate of polar wander (We denoted  the rate of TPW by 
 $\dot{W} = \vert \dot{\mitbf{\omega}} \vert$) .
Evaluating the scalar product $\dot{W}^2 = \dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}}$ gives
\begin{equation}
\dot{\Theta}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
\label{eq:polar_wander_rate_tensor}
\end{equation}
The scalar terms on the right-hand-side of Equation (14) are now independent of the choice of coordinate system as long as we are consistent.
Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$. In this case the orientation of 
$\mitbf{\omega}$ is defined in the $\mathbf{E}$-frame using colatitude $\theta$ and longitude $\phi$ (see Figure~\ref{fig:reference_frames}).
Plugging this description of $\mitbf{\omega}$ into Equation~\eqref{eq:polar_wander_rate_tensor},
and after some tedious algebra, we find

\begin{equation}
\begin{aligned}
\dot{W}^2 = 
&\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
\label{eq:polar_wander_rate}
\end{aligned}
\end{equation}
This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}. Once the orientation of $\mitbf{\omega}$ is specified in the E-frame by angles $\theta$ and $\phi$, we can evaluate the rate of TPW in terms of the principal moments of $\mathbf{E}$.
Two special cases of this equation are worth  noting. First, if $\mitbf{\omega}$  coincides with  the normal to the octahedral plane
(i.e. the plane defined by a normal with direction cosines all equal to $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
then it can be expressed in terms of the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator 
($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
\begin{equation}
\begin{aligned}
\dot{W}^2
&= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2}. \\
\end{aligned}
\end{equation}
The second invariant of the stress deviator is commonly used in the theories of plasticity
to characterize the state of shear stress.
The second invariant of the moment of inertia deviator can be viewed as a ``rotational stress" that drives a TPW response.

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/reference_frames.pdf}
\caption{Relevant vectors and angles for the TPW analysis. The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively.  The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $\mathbf{e}_3$-axis. For illustration, the longitude $\phi$ of the rotation axis is taken to be zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane. In general $\phi$ is not zero. True polar wander moves the rotation axis towards the $\mathbf{e}_3$-axis. Observations of $\dot{\mitbf{\omega}}$ are made in the geographic frame, which drifts relative to the E-frame.  }
\label{fig:reference_frames}
\end{figure*}

The second special case occurs when the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_3$ plane ($\phi = 0$).
Equation~\eqref{eq:polar_wander_rate} reduces to
\begin{equation}
\vert \dot{W} \vert = 
\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1)\, .
\label{eq:simple_milankovitch}
\end{equation}
An equivalent expression was given by \citet{tsai2007theoretical}. We adopt Equation (17) as the basis for our dimensional analysis. Note that
the maximum  rate is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).

From Equation~\eqref{eq:simple_milankovitch} it is clear that the important quantities for estimating the rate of true polar wander ($\dot{W}$)
are $\theta$ and $\lambda_3 - \lambda_1$. Both of these quantities depend on the structure and dynamics of mantle convection.  The size of
convective anomalies directly determines $\mathbf{E}$, and hence $\lambda_3-\lambda_1$. We also expect convection to alter the orientation of the principal axes of $\mathbf{E}$, which drives a change in $\theta$. Relaxation of the rotation axis back toward the principal axis of $\mathbf{E}$ acts to reduce $\theta$. In the next section we focus on the role of mantle convection in setting $\lambda_3-\lambda_1$ and in driving changes in $\theta$.
